Lie Algebras and Groups\vspace{-1ex}

### 2019-01-11 Hopf Algebra Notes/Hopf Algebras.tex
Update Notes

%Hopf algebra stuff
\newcommand*{\Vectk}{\operatorname{Vect}_k}
\newcommand*{\Algk}{\operatorname{Alg}_k}
\newcommand*{\Coalgk}{\operatorname{Coalg}_k}

@@

>>>>

\DeclareMathOperator{\Hom}{Hom}

%Hopf algebra stuff
\newcommand*{\Vectk}{\operatorname{Vect}_k}
\newcommand*{\Algk}{\operatorname{Alg}_k}
\newcommand*{\Coalgk}{\operatorname{Coalg}_k}
\DeclareMathOperator{\Coalg}{Coalg}
\DeclareMathOperator{\Alg}{Alg}
\newcommand*{\scrT}{\ensuremath\mathscr{T}}

@@

\section{January 11, 2019}
The plan for today is to talk about:
\begin{itemize}
	\item Convolution Algebras
	\item Antipodes
	\item Duality
	\item (Co-)Modules
\end{itemize}

\subsection{Convolution Algebras}
 Let $\scrT$ be a symmetric tensor category. We can usually think of $\scrT=\Vectk$, but 
 there is a problem since $\Vectk$ is equivalent to the category of Hopf algebras over $k$,
 while this is not generally true.

 We also need that $\scrT$ is $k$-linear (that is, enriched as a category over $k$). This
 means that $\Hom_\scrT (A,B)\in\Vectk$.

 \begin{thm}
	Let $\scrT$ be as above. Then $\Hom_\scrT(C,A)$ is an algebra and $\Hom_\scrT(-,-):(\Coalg_\scrT)^{op}\times\Alg_\scrT\to\Alg_k$
	is a functor.
 \end{thm}
 \begin{prf}
	Let $A$ be an algebra object in $\scrT$ and $C$ be a coalgebra object in $\scrT$. Then $1_{\Hom} :=u_A\circ\varepsilon_C:C\to 1_\scrT$ and define
	\[f\ast g := m_A(f\otimes g)\Delta_C:C\to A.\]

	Then using Lemma \ref{lem-func} and the fact that $A$ and $C$ are (co)algebra objects,
	we can see that the product $\ast$ satisfies the axioms required.
 \end{prf}
 \begin{lem}
	\begin{enumerate}
		\item $1_{\Hom}\ast f = m_A(u\otimes 1)(1\otimes f)(\varepsilon\otimes 1)\Delta.$
		\item $f\ast 1_{\Hom}=m_A(1\otimes u)(f\otimes 1)(1\otimes \varepsilon)$
		\item $(f\ast g)\ast h=m_A(m_A\otimes 1)((f\otimes g)\otimes h)(\Delta_C\otimes 1)\Delta_C$
		\item $f\ast(g\ast h)=m_A(1\otimes m_A)(f\otimes(g\otimes h))(1\otimes \Delta)\Delta$
	\end{enumerate}
	\label{lem-func}
 \end{lem}

 \begin{prf}
	\subsubsection*{(a)}
	\begin{align*}
		1_{\Hom}\ast f&= m_A(1_{\Hom}\otimes f)\Delta_C\\
		&=m_A(u_A\circ\varepsilon\otimes f)\Delta_C\\
		&=m_A(u\otimes 1)(\varepsilon\otimes 1)(1\otimes f)\Delta
	\end{align*}
	\subsubsection*{(b)}
	Same as (a), essentially.
	\subsubsection*{(c) and (d)}
	\begin{align*}
		(f\ast g)\ast h&= m_A((f\ast g)\otimes h)\Delta\\
		&=m_A((m_A(f\otimes g)\Delta)\otimes h)\Delta\\
		&=m_A(m_A\otimes 1)((f\otimes g)\otimes h)(\Delta\otimes 1)\Delta
	\end{align*}
	and the other is analogous.
 \end{prf}

 \begin{defn}
	$V\in\scrT$ is a \textbf{Hopf algebra object} if:
	\begin{itemize}
		\item $V$ is a bialgebra object in $\scrT$ and
		\item There is a map $S:V\to V$ that is $(\id_V)^{-1}$ with respect to $\ast$.
	\end{itemize}
 \end{defn}
 \begin{rmk}
	Notice here that $\id_V\in\Hom_\scrT(V,V)$, the identity map in $\scrT$. We are \textit{not}
	taking about $1_{\Hom}=u\circ\varepsilon$.

	Also, we call $S$ an \textbf{antipode.}
 \end{rmk}

 \subsection{Duality}
 Notice that when $C=1_\scrT$ (that is the tensor identity), $\Hom_\scrT(1_\scrT,-):\Alg_\scrT\to\Algk$
 is a functor. Same for the dual from $\Coalg_\scrT$. This second one gives us a chance to talk about duality.

 \begin{lem}
	Let $\scrT$ be the category of finite dimensional vector spaces over $k$.
	Then $(-)^*:\scrT\to\scrT^{op}$ is an equivalence.
 \end{lem}
 This uses $(V\otimes W)^*=W^*\otimes V^*$.
 \begin{cor}
	$V$ is an algebra over $k$ $\Leftrightarrow$ $V^*$ is a coalgebra over $k$.
	And vice versa.
 \end{cor}

 Recall $S=(\id_V)^{-1}$. Thus
 \[S\ast \id_V=1_{\Hom}=\id_V\ast S\]
 The diagram we have here is
 \begin{figure}[h]
	 \centering
	 \begin{tikzcd}
		V\otimes V\ar[rr,"S\otimes \id_V"] & & V\otimes V\ar[d,"m_V"]\\
		V\ar[u,"\Delta"]\ar[d,"\Delta"]\ar[r,"\varepsilon_V"] & k \ar[r,"u_V"] & V\\
		V\otimes V\ar[rr,"\id_V\otimes S"] & & V\otimes V\ar[u,"m_V"]
	 \end{tikzcd}
 \end{figure}

 \subsection*{Modules/Comodules}
 \begin{defn}
	Let $A$ be an algebra object in $\scrT$. A \textbf{left $A$ module} is $M\in\scrT$
	with a morphism
	\[m_M:A\otimes M\to M\]
	such that the diagrams in Figure \ref{fig-mod} commute.
\end{defn}
\begin{figure}[h]\label{fig-mod}
	\centering
	\begin{tikzcd}
		A\otimes A\otimes M\ar[r,"m_A\otimes 1"]\ar[d,"1\otimes m_M"] & A\otimes M\ar[d,"m_M"]\\
		A\otimes M\ar[r,"m_M"] & M
	\end{tikzcd}
	\begin{tikzcd}
		A\otimes M\ar[r,"m_M"] & M\\
		1_\scrT\otimes M\ar[ur,"\sim"]\ar[u,"u_A\otimes 1"] &
	\end{tikzcd}
	\caption{Module diagrams}
\end{figure}
\begin{rmk}
	Note that we don't necessarily need that $M$ lie in $\scrT$. We could instead just 
	rely on an algebra homomorphism $\varphi:A\to \Hom_{\scrT}(M,M)$ and proceed as usual.
\end{rmk}

### 2019-01-11 Lie Algebra Notes/Lie Algebras.tex
Update Notes

%set up theorem/definition/etc envs
%Problems will be created using their own counter and style
\theoreminframepreskip{0pt}
\theoreminframepostskip{0pt}
\newframedtheorem{prob}{Problem}[part]
\renewcommand\theprob{\arabic{part}.\arabic{prob}}

@@

%Create a new env that references a theorem and creates a 'primed' version
%Note this can be used recursively to get double, triple, etc primes
\newenvironment{thm-prime}[1]
  {\renewcommand{\thethm}{\ref{#1}$'$}%
   \addtocounter{thm}{-1}%
   \begin{thm}}
  {\end{thm}}

@@

%Man, that's really good! Let's use the same thing for definitons.
\newenvironment{def-prime}[1]
  {\renewcommand{\thethm}{\ref{#1}$'$}%
   \addtocounter{thm}{-1}%
   \begin{def}}
  {\end{def}}

@@

%Lie algebra stuff
\DeclareMathOperator{\gl}{\mathfrak{gl}}
\let\sl\relax
\DeclareMathOperator{\sl}{\mathfrak{sl}}
\DeclareMathOperator{\so}{\mathfrak{so}}
\let\sp\relax
\DeclareMathOperator{\sp}{\mathfrak{sp}}
\DeclareMathOperator{\tr}{tr}
\DeclareMathOperator{\ad}{ad}

@@

\begin{lem}
	For all \textbf{complex} semisimple Lie algebras $L$ it turns out that the group
	generated by $\exp\ad x$ ($\ad x(y)=[xy]$) coincides with the group generated by all 
	nilpotent $\ad x$.
\end{lem}
\subsection{Adjoint group}
The last thing for today is to define the adjoint group:
\begin{defn}
	Let $L$ be a Lie algebra, then 
	\[\operatorname{Int}(L)=\exp\ad L\]
	is the \textbf{Adjoint group} of $L$. It is a subgroup (so we believe) of the Lie
	Group associated to $L$.
\end{defn}

@@

\end{document}

>>>>

%set up theorem/definition/etc envs
%Problems will be created using their own counter and style
\theoreminframepreskip{0pt}
\theoreminframepostskip{0pt}
\newframedtheorem{prob}{Problem}[section]
\newenvironment{hwprob}[1]
{\renewcommand{\theprob}{#1}%
 \addtocounter{thm}{-1}%
 \begin{prob}}
{\end{prob}}

@@

%Create a new env that references a theorem and creates a 'primed' version
%Note this can be used recursively to get double, triple, etc primes
\newenvironment{thmprime}[1]
  {\renewcommand{\thethm}{\ref{#1}$'$}%
   \addtocounter{thm}{-1}%
   \begin{thm}}
  {\end{thm}}

@@

%Man, that's really good! Let's use the same thing for definitons.
\newenvironment{defprime}[1]
  {\renewcommand{\thethm}{\ref{#1}$'$}%
   \addtocounter{thm}{-1}%
   \begin{defn}}
  {\end{defn}}

@@

%some nice shortcuts
\DeclareMathOperator{\calB}{\mathcal{B}}

%Lie algebra stuff
\DeclareMathOperator{\gl}{\mathfrak{gl}}
\let\sl\relax
\DeclareMathOperator{\sl}{\mathfrak{sl}}
\DeclareMathOperator{\so}{\mathfrak{so}}
\let\sp\relax
\DeclareMathOperator{\sp}{\mathfrak{sp}}
\DeclareMathOperator{\tr}{tr}
\DeclareMathOperator{\ad}{ad}
\DeclareMathOperator{\Der}{Der}
\DeclareMathOperator{\Int}{Int}
\DeclareMathOperator{\Rad}{Rad}
\let\t\relax
\DeclareMathOperator{\t}{\mathfrak{t}}
\DeclareMathOperator{\n}{\mathfrak{n}}
\DeclareMathOperator{\diag}{diag}

@@

\begin{lem}
	For all \textbf{complex} semisimple Lie algebras $L$ it turns out that the group
	generated by $\exp\ad x$ ($\ad x(y)=[xy]$) coincides with the group generated by all 
	nilpotent $\ad x$.
\end{lem}
\subsection{Adjoint group}
The last thing for today is to define the adjoint group:
\begin{defn}
	Let $L$ be a Lie algebra, then 
	\[\operatorname{Int}(L)=\exp\ad L\]
	is the \textbf{Adjoint group} of $L$. It is a subgroup (so we believe) of the Lie
	Group associated to $L$.
\end{defn}
\begin{rmk}
	Actually after talking to Monty, $\exp\ad L$ \textit{is} (essentially) the Lie group
	corresponding to $L$. Such a group is not unique, however.
\end{rmk}

@@

%%%% HW 1 - Due Jan 18 %%%%
\newpage

\section*{HW 1 -- Due January 18}
Do problems 1.9, 2.1, 3.8. 3.9, and 4.3.

\begin{hwprob}{1.9}
	When $\ch k=0$, show that each classical Lie algebra $A_l,B_l, C_l$ and $D_l$ satisfies $[LL]=L$. (This shows again that each algebra consists of trace zero matrices.)
\end{hwprob}
\begin{sol}
	\subsubsection*{Type $A_k$:}
	Recall that a basis for $A_l$ is the collection $\mathcal{B}=\{e_{ij}|i\ne j\}\cup\{e_{ii}-e_{i+1,i+1}|1\le i\le l\}$. Using linearity of the bracket,
	it then suffices to check that $\calB$ is contained in the algebra generated by $[\calB \calB]$. To see this, we use that
	\[[e_{ij}e_{kl}]=\delta_{jk}e_{il}-\delta_{li}e_{kj}.\]

	First, notice we have
	\[[e_{i,i+1}e_{i+1,i}]=e_{ii}-e_{i+1,i+1},\]
	so it suffices to show we also have the $e_{ij}$ for $i\ne j$.

\end{sol}

\begin{hwprob}{2.1}
	Prove that the set of all inner derivations $\ad x$ for $x\in L$ is an ideal of $\Der L$.
\end{hwprob}
\begin{sol}

\end{sol}

\begin{hwprob}{3.8}
	Let $L$ be nilpotent. Prove that $L$ has an ideal of codimension 1.
\end{hwprob}
\begin{sol}

\end{sol}

\begin{hwprob}{3.9}
	Prove that every nilpotent Lie algebra L has an outer derivation (see 1.3) as follows: Write $L=K+Fx$
	for some ideal $K$ of codimension 1. Then $C_L(K)\ne 0$ (why?). Choose $n$ so that $C_L(K)\subseteq L^n$,
	$C_L(K)\not\subseteq L^{n+1}$ and let $z\in C_L(K)-L^{n+1}$. Then the linear map $\delta$ sending $K$ to zero
	and $x$ to $z$ is an outer derivation.
\end{hwprob}
\begin{sol}

\end{sol}

\begin{hwprob}{4.3}
	This exercise illustrates the failure of Lie's theorem when $F$ is allowed to have positive characteristic.
	Consider the $p\times p$ matrices
	\[x=\begin{pmatrix}
		0 & 1 & 0 & \cdots & 0\\
		0 & 0 & 1 & \cdots & 0\\
		\vdots & \vdots & \vdots & \ddots & \vdots\\
		0 & 0 & 0 & \cdots & 1\\
		1 & 0 & 0 & \cdots & 0
	\end{pmatrix}\quad y=\diag(0,1,\dots, p-1)\]
	Check that $[xy]=x$, hence that $x$ and $y$ span a two dimensional solvable subalgebra $L$ of $\gl(p,F)$. Verify that $x$ and $y$ have no common eigenvector.
\end{hwprob}
\begin{sol}

\end{sol}
\newpage
%%%% End HW 1 %%%%

\section{January 11, 2019}
A few more things about $\Int L$ in classical cases:

Begin with $n\times n$ matrices $M$ over $\R$ or $\C$. Given any such $M$, we have
\[e^M=\sum_{i>0}\frac{M^i}{i!}\]
always converges (that is, the series for each entry converges). Futhermore $e^Me^{-M}=I$.
\begin{rmk}
	Notice that over \textit{any} field over characteristic zero whenever $M$ is nilpotent 
	this still holds.
\end{rmk}

Now if $M$ is skew-adjoint with respect to a bilinear form $(\cdot,\cdot)$, then 
\[(M^iv,w)=(-1)^i(v,M^iw)\]
whence
\[(e^Mv, w)=(v,e^{-M}w)\]
and so 
\[(e^Mv,e^Mw)=(v,w)\]
for all $v,w\in\R^n$ or $\C^n$.

But notice that then $e^M$ is an \textbf{isometry of} or \textbf{preserves} the form. Moreover, 
if we set $A=\ad M$, then $A$ is given as the difference of two \textit{commuting} linear maps, 
left and right multiplication by $M$.

Using the binomial theorem: $e^AN=e^MNe^{-M}$. This, in a nutshell, is why the formulas for
$\Int L$ all involved modding out by the center, since conjugation by a scalar matrix is trivial! :)

\subsection{Combining Ideals}
Now we return to the case of general Lie algebras over arbitrary fields (no longer assuming $\ch k=0$). 
We have seen that a subspace $I\le L$ is an ideal if $[IL]\subseteq I$.

If $I,J\lhd L$, then $I+J$ is an ideal, as well as
\[[IJ]=\{\sum_1^n [x_iy_i]|x_i\in I, y_i\in J\}.\]

We also are sometimes interested in the direct sum of ideals $\oplus L_i$ where $[L_iL_j]=0$ when $i\ne j$.

\subsection{Examples}
$0$ and $L$ are ideals as well as the \textbf{center} $Z(L)$.

\subsection{Characterizing Lie Algebras as Linear}
We now want to realize many Lie algebras as (isomorphic to) linear ones. For now, note that the adjoint map $\ad$
is a Lie algebra homomorphism by the Jacobi identity. Then $\ker\ad=Z(L)$, so when $Z=0$, then $L$ is isomorphic to 
a linear Lie algebra.

\begin{defn}
	If $Z(L)$, and the map $\ad:L\to L$ is an isomorphism, we say $L$ \textbf{acts faithfully}
	on itself (via $\ad$).
\end{defn}

\begin{defn}
	If we have a homomorphism $L\to\gl(n,k)$, then we call $k^n$ an $L$-\textbf{module}.
	We define the action of $L$ on $K^n$ through the homomorphism.
\end{defn}

Notice that $L$ is always an $L$ module over itself.

One can check, using the realization of $\sl(n,k)$ we saw last time that (when $\ch k\nmid n$)
that $\sl(n,k)$ is simple. This is actually the Lie algebra analogue of the fact that $M_n(K)$ is simple
as a ring over any field $K$.

\begin{rmk}
	Notice that if $\ch k |n$$,$ then \textit{any} scalar matrix $aI\in\sl(n,k)$
	so $Z(\sl(n,k))$ is nontrivial.
\end{rmk}

In any event you can look up the proof for $\sl(2,k)$ in the book. The thing to glean here is that
\[[xy]=h,\quad [hx]=2x,\quad [hy]=-2y\]
for $h=e_{11}-e_{22}$, $x=e_{12}$ and $y=e_{21}$. This will keep popping up throughout the course.

\subsection{Ideals of Linear Lie Algebras}
Many subalgebras of $\sl(n,k)$ or $\gl(n,k)$ have many ideals (which happen to be ring-theoretic ideals).

\begin{defn}
	Some notation:
	\[\t(n,k)=\{\text{upper-triangular matrices}\}\] 
	also 
	\[\n(n,k)=\{\text{strictly-upper-triangular matrices}\}\]
\end{defn}

\subsection{Derived Series}
For any Lie algebra, we can define
\begin{defn}
	The \textbf{derived series} of a Lie algebra $L$ is defined to be
	\[L^{(0)}=L,\quad L^{(1)}=[LL],\quad L^{(n+1)}=[L^{(n)}L^{(n)}]\]
	which form an ascending chain of ideals.
\end{defn}

\begin{defn}
	If $L^{(n)}$ for some $n$, and thus for all sufficiently large $n$, we call $L$ \textbf{solvable.}
\end{defn}
\begin{rmk}
	Note that actually there is a cooresponding definition for groups using commutators $[G,G]$.
	This was defined first and ported to Lie algebras. It actually happened in the reverse direction for:
\end{rmk}

\begin{defn}
	The \textbf{central series} of $L$ is 
	\[L^0=L,\quad L^1=[LL],\quad L^2=[L,L^1],\dots\]
	and if $L^n=0$ for some $n$, then we call $L$ \textbf{nilpotent.}
\end{defn}
\begin{rmk}
	Equivalently, we can say that $L$ is nilpotent if \textit{any $n$-fold bracket} in $L$ is zero.
\end{rmk}
\begin{rmk}
	Notice that every nilpotent Lie group is solvable, but the converse fails: $t(n)$ is solvable by not nilpotent.
\end{rmk}

\subsection{Examples of solvable algebras}
One can easily check that every ideal and homomorphic image of a solvable algebra is solvable. Therefore if $I$ and $J$ are solvable ideals of $L$, then 
so is $(I+J)/J\cong I/(I\cap J)$ is solvable, as is $I+J$. Furthermore if $I$ and $L/I$ are solvable, then $L$ is.

\begin{defn}
	$\Rad L$, the \textbf{radical of $L$}, is the unique largest solvable ideal of $L$.
\end{defn}
\begin{rmk}
	Any finite dimensional algebra has a radical (taken to be the sum of all solvable ideals in $L$).
\end{rmk}

Using this definition, we are able to give two different characterizations:
\begin{defn}[Semisimple Lie Algebra]\label{def-semisimple}
	$L$ is semisimple if $\Rad L=0$.
\end{defn}
\begin{defprime}{def-semisimple}[Semisimple Lie Algebra]
	$L$ is semisimple if it is a direct sum of simple Lie algebras.
\end{defprime}
This equivalence will not become clear until we learn about the Killing form, but at least it jives
with our experience with representation theory and semisimple modules.
\end{document}
